Mixed effect tweedie model in TMB

### 2020-06-19 analysis/tmb-sensor-explore/3-fit-do.Rmd
fitting climate models

---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Fit spatial temporal models to bottom temperature

@@

glimpse(d_trawl)
```

@@

```{r}
# # filter surveys if desired
# d_trawl <- d_trawl %>% 
  # filter(ssid != 1) %>%
  # filter(ssid != 3) %>%
  # filter(ssid != 4) %>%
  # filter(ssid != 16) %>%
  # remove obvious sensor fails, d_trawl %>% filter(depth_max<10), fishing event ids = 481861, 2179089
  # filter(depth_m_max>10) %>% 
```

```{r}

d_trawl1 <- d_trawl %>% filter(!(year == 2016 & ssid ==4)) #%>% filter(ssid != 16) %>% filter(ssid != 4)    
#test <- d_trawl1 %>% filter(year == 2016 & ssid ==16) 
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 800) ### TODO: rerun with 500
plot_spde(spde)
```

Test temp model on this subset of data
DOY doesn't improve model
```{r eval=FALSE}
    starttime1 <- Sys.time()

    temp_model_DOY <- sdmTMB::sdmTMB(d_trawl1,
      temp_scaled ~ 0 + as.factor(year), # + DOY_scaled,
      time_varying = ~ 0 + depth_scaled + depth_scaled2,
      time = "year", spde = spde,
      family = gaussian(link = "identity"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = FALSE,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

saveRDS(temp_model_all, file = (here::here("analysis/tmb-sensor-explore/models/model-temp-2008-on.rds")))

# saveRDS(temp_model_DOY, file = (here::here("analysis/tmb-sensor-explore/models/model-temp-2008-on-DOY.rds")))
# time1
# temp_model_DOY
temp_model_all
```

```{r eval=FALSE}
get_diag(temp_model_all, response = "temp_scaled", variable = "depth_scaled")
```

Add temperature predictions to newdata grid for DO predictions
```{r eval=FALSE}
temperature <- readRDS(here::here("analysis/VOCC/data/predicted_temp_allyears_revised2"))
years <- unique(d_trawl1$year) 
temperature <- filter(temperature, year %in% years)
unique(temperature$year)

temperature$temp <- temperature$est
#temperature$temp_scaled2 <- temperature$est^2
temperature$raw_depth <- temperature$depth
temperature$depth <- log(temperature$depth)
temperature$DOY_scaled <- 0

newdata <- temperature %>% 
  select(-est, - est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st, -temperature_c) %>%
  scale_predictors(predictors = c(quo(depth),quo(temp))) %>% 
  mutate(depth = raw_depth)

glimpse(newdata)
```

Current best model of DO
```{r eval=FALSE}
    starttime1 <- Sys.time()

    do_model_all_DOY <- sdmTMB::sdmTMB(
      log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
      data = d_trawl1,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + depth_scaled3,
      time = "year", 
      spde = spde,
      family = gaussian(link = "identity"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = FALSE,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

saveRDS(do_model_all_DOY, file = (here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds")))
time1
do_model_all_DOY
```

```{r eval=FALSE}
# do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-all-DOY.rds"))
# do_model_DOY

get_diag(do_model_all_DOY, response = "log_do", variable = "depth_scaled")

# get_diag(do_model_DOY, response = "log_do", variable = "temp_scaled")
```

Other models tried:
```{r eval=FALSE}
do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds"))
do_model_DOY
#p_nd <- predict(do_model_DOY, newdata = newdata)
get_diag(do_model_DOY, response = "log_do", variable = "depth_scaled")
```

```{r eval=FALSE}
#with both DOY and depth as time-varying random effects
do_model_re_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-re-DOY.rds"))
do_model_re_DOY
get_diag(do_model_re_DOY, response = "log_do", variable = "depth_scaled")
```

```{r eval=FALSE}
do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-DOY.rds"))
do_model_DOY
#p_nd <- predict(do_model_DOY, newdata = newdata)
get_diag(do_model_DOY, response = "log_do", variable = "depth_scaled")

```{r eval=FALSE}
do_model_depth2 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-depth2.rds"))
do_model_depth2
get_diag(do_model_depth2,response = "log_do", variable = "depth_scaled")
#p_nd <- predict(do_model_depth2, newdata = newdata)
```

```{r eval = FALSE}
do_model_depth3 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-depth3.rds"))
get_diag(do_model_depth3,response = "log_do")
#p_nd <- predict(do_model_depth3, newdata = newdata)
```

```{r eval=FALSE}
do_model_all <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-all.rds"))
#p_nd <- predict(do_model_all, newdata = newdata)
do_model_all
get_diag(do_model_all, response = "log_do", variable = "depth_scaled")
```

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r eval=FALSE}
do_model_all <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-poly.rds"))
p_nd <- predict(do_model_all_DOY, newdata = newdata)
p_nd$do_est <- exp(p_nd$est)

p_nd <- p_nd %>% filter(depth>15)
hist(p_nd$do_est)

max(p_nd$do_est)
max(d_trawl1$do_mlpl)
glimpse(p_nd)
min(p_nd$raw_depth)
min(d_trawl1$raw_depth)
p_nd[p_nd$do_est>8, ]

p_nd <- p_nd %>% filter(depth>15)
p_nd_filtered[p_nd_filtered$do_est>8, ]

saveRDS(p_nd, file = (here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))
```

Save new predictions without QCS gap and WCVI 2016 DO
```{r eval=FALSE}
#new temp and do prediction grid for vectors

# model1 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-temp-2008-on-used.rds")
#model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-all-DOY-used.rds")

# fix RW additon error
model1 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-500kn.rds"))
# update model to exclude WCVI 2016 outliers
model2 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds"))

@@

# get scaling used in model
d <- all_sensor %>%
  # filter missing do data 
  dplyr::filter(!is.na(do_mlpl)) %>%
  # remove 2007 data because clearly faulty or scaled differently
  dplyr::filter(year>2007)
d$temp_scaled <- scale(d$temperature_c)
attributes(d$temp_scaled)

predtemp <- pred_temp %>% mutate(temp_scaled = (est - 6.49 )/1.296, temp_scaled2 = ((est - 6.49 )/1.296)^2, depth_scaled3 = depth_scaled^3) %>% select(-est, -est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st)
predtemp <- predtemp %>% filter(year > 2007) 
predtemp$DOY_scaled <- 0

pred_do <- predict(model2, newdata = predtemp)

pred_do$temp <- pred_do$temp_scaled*1.296 + 6.49

pred_do$log_do <- pred_do$est
pred_do$do_est <- exp(pred_do$log_do)

@@

saveRDS(pred_do, file=here::here("analysis/VOCC/data/predicted-DO-june-2020.rds"))

Plot predictions from saved DO predictions file

```{r}
#p_nd <- readRDS((here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))
# p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-new.rds")))

p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-june-2020.rds")))

@@

```{r}
p_nd1 <- p_nd %>% filter(ssid!=16) %>% filter(ssid!=4)

breaks <- scales::trans_breaks(no_trans[["transform"]], no_trans[["inverse"]],n = 6)
labels <- function(x) format(x, digits = 2, scientific = FALSE)

@@

```{r, eval=FALSE}
#glimpse(p_nd)

omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
   scale_fill_gradient2( high = "Steel Blue 4",
  low = "Red 3", limits = c(-0.65,1))+
  labs(fill = " ") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.82,.75))

@@

```{r}
png(
  file = "figs/do-predictions-without-wcvi2016.png", 
  res = 600,
  units = "in",
  width = 11, 
  height = 6.5
)

gridExtra::grid.arrange(
  grobs = list(pred_1, st_1n3, omega_s_all, pred_4, st_4, st_16 ),
  widths = c(2, 2, 1.75),
  heights = c(2, 1.5),
  layout_matrix = rbind(c(1, 2, 3),
                        c(4, 5, 6)),
  top = grid::textGrob("Dissolved oxygen estimates from sdmTMB model")
)

# gridExtra::grid.arrange( nrow = 2, heights = c(2,1.5), 
#   top = grid::textGrob("Dissolved oxygen estimates from sdmTMB model")) 
dev.off()
```

```{r}
#do_cog <- get_cog(p_nd)

#temp_cog %>%
#  reshape2::dcast(year ~ coord, value.var = "est") %>%
#  ggplot(aes(X, Y, colour = year)) + geom_path(arrow = arrow()) +
#  scale_color_viridis_c()
```

```{r}
#temp_cog %>%
#  ggplot(aes(year, est, ymin = lwr, ymax = upr)) +
#  geom_line() +
#  geom_ribbon(alpha = 0.2) +
#  facet_wrap(~coord, scales = "free_y")
```

>>>>

---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Fit spatial temporal models of bottom DO

@@

# glimpse(d_trawl)
```

Ranges sampled
```{r}
paste("depth") 
range(d_trawl$raw_depth)
paste("DO") 
range(d_trawl$do_mlpl)
```

@@

Remove WCVI in 2016 due to extreme values and make 500 knot mesh
```{r}
d_trawl1 <- d_trawl %>% filter(!(year == 2016 & ssid ==4))

spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 500) 
plot_spde(spde)
```

Current best model of DO
```{r }
do_model_all_DOY <- sdmTMB::sdmTMB(
      log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
      data = d_trawl1,
      time_varying = ~ 0 + depth_scaled + depth_scaled2, 
      time = "year", 
      spde = spde,
      family = gaussian(link = "identity"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = FALSE,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )

saveRDS(do_model_all_DOY, file = (here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds")))
```

```{r eval=FALSE}
get_diag(do_model_all_DOY, response = "log_do", variable = "depth_scaled")
```

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r eval=FALSE}
model1 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-500kn.rds"))
model2 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-depth2.rds"))

@@

# get scaling used in model
sensor_temp_scaled <- scale(d_trawl$temp)
attributes(sensor_temp_scaled)

predtemp <- pred_temp %>% mutate(
  temp_scaled = (est - attributes(sensor_temp_scaled)[[2]] )/attributes(sensor_temp_scaled)[[3]], 
  temp_scaled2 = (temp_scaled)^2) %>% select(-est, -est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st)
predtemp <- predtemp %>% filter(year > 2007) 
predtemp$DOY_scaled <- 0

# make DO predictions
pred_do <- predict(model2, newdata = predtemp)

# rename and back transform climate variables
pred_do$temp <- pred_do$temp_scaled * attributes(sensor_temp_scaled)[[3]] + attributes(sensor_temp_scaled)[[2]]
pred_do$log_do <- pred_do$est
pred_do$do_est <- exp(pred_do$log_do)

@@

saveRDS(pred_do, file=here::here("analysis/VOCC/data/predicted-DO-depth2.rds"))
```

Plot predictions from saved DO predictions file
```{r}
p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-depth2.rds")))

@@

Plot each survey separately
```{r}
p_nd1 <- p_nd %>% filter(ssid!=16) %>% filter(ssid!=4)

breaks <- scales::trans_breaks(no_trans[["transform"]], no_trans[["inverse"]], n = 6)
labels <- function(x) format(x, digits = 2, scientific = FALSE)

@@

```{r, eval=FALSE}
omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
   scale_fill_gradient2( high = "Steel Blue 4",
  low = "Red 3", limits = c(-0.65,1))+
  labs(fill = " ") + 
gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.82,.75))

@@

```{r, eval= F}
png(
  file = "figs/do-predictions-without-wcvi2016.png", 
  res = 600,
  units = "in",
  width = 11, 
  height = 6.5
)

gridExtra::grid.arrange(
  grobs = list(pred_1, st_1n3, omega_s_all, pred_4, st_4, st_16 ),
  widths = c(2, 2, 1.75),
  heights = c(2, 1.5),
  layout_matrix = rbind(c(1, 2, 3),
                        c(4, 5, 6)),
  top = grid::textGrob("Dissolved oxygen estimates from sdmTMB model")
)
dev.off()
```
